Parity games in pushdown systems with pebbles

### 2023-06-11 reachability/article.tex
up

We consider here a different approach to extending pushdown systems, by allowing them to test the stack content against precise values.
Pushdown systems which can test the stack content against regular-expressions are
equivalent to pushdown systems,
so pushdown systems which can test the stack content against specific words are too.
We consider instead checks against unspecified word values that can be instantiated one after the other,
providing the pushdown systems the ability to
store information on some stack contents for later comparisons.

@@

\subsection{Parity Games}

We first recall the definition of  two  player parity  games.
Two players, player $0$ and player $1$, move along an arena $\mathcal{G}=(V_0,V_1,E,\Omega)$ which is a labelled graph, composed of
two  disjoint  sets of vertices, $V_0$ and $V_1$ respectively associated with player $0$ and player $1$, 
a set of edges $E \subseteq V \times V$, where $V=V_0 \uplus V_1$,
% We require that the out-degree of each vertex is at least one. This ensures that every finite path in (V, E) can be prolonged.
and a mapping $\Omega:V \to \{0,1, \ldots, m\}$,
$m < \omega$, which
assigns a priority to each vertex. %Then the {\em initialized game} $(\mathcal{G}, v_I)$ is given with an initial vertex $v_I \in V$.

A {\em partial play} of $\mathcal{G}$ is a sequence of vertices $v_0 v_1 \ldots \in V^\infty$ connected by the transition relation, i.e. such that $(v_i, v_{i+1}) \in E$.

A {\em play} of $\mathcal{G}$ is a partial play that is {\em maximal} in the following sense: if it is finite, then the last vertex $v_k$ is such that for all $v' \in V$, $(v_k, v') \notin E$. Else, the play is an infinite sequence $ \pi  = v_0 v_1 \ldots \in  V^{\omega} $.

We consider min-parity games: in order to find out who wins a play, we need to take a
look at the positions which are reached infinitely often and compute their priorities using 
$\Omega$.
Player $0$ wins the play $\pi$  iff  $\min( \{ \Omega(v) \mid v \in \text{\textit{Inf}}( \pi) \})$ is even, else, player $1$ wins. The winner of a finite play is the player whose opponent is unable to move, i.e. the winner of a play such that the last vertex of the play is in $V_{0}$ (resp. $V_1$) is player $1$ (resp. player $0$).

A {\em strategy} $\sigma_i : V^* V_i \to V$ for player $i \in \{ 0,1\}$ is a function that, given a sequence of vertices $ w v$  with $w\in V^*$ and $v \in V_i$,
provides, if possible, a successor $v'$ such that $(v,v') \in E$.
A strategy is called {\em memoryless} if its output only depends on the final vertex of the sequence $v \in V_i$, i.e. if there exists a function $f_i : V_i \to V$ such
that for all $w\in V^*$, $v \in V_i$, we have $\sigma_i(wv) = f_i(v)$.

A partial play $\pi$ is {\em consistent} with a strategy $\sigma_i$ if sequences of vertices that end in $V_i$ along this play all have successors according to the strategy. Given a starting vertex $v_I$,   a strategy $\sigma_0$ for player $0$ and a strategy $\sigma_1$ for player $1$, the play
starting with $v_I$ that is consistent with both strategies is unique and is called the {\em resulting play} of $\sigma_0$ and $\sigma_1$ starting from $v_I$. It is 
denoted by $\pi(v_I, \sigma_0, \sigma_1)$.

A strategy $\sigma_i$ for player $i$ from position $v_I \in V$ is a {\em winning strategy from $v_I$} if 
no matter which
strategy $\sigma_{1-i}$ the other player pursues,
% the resulting play  starting from $v_I$  
%  results in 
% player $i$ winning.
player $i$ wins the play $\pi(v_I, \sigma_0, \sigma_1)$.

The above definitions are essentially the same for finite and infinite arenas.
%
%
In this paper we are going to concern ourselves with parity games over potentially infinitely large arenas, \textcolor{black}{thus we define the following decision problem in a general abstract manner. In particular, we do not discuss representation of the input}.

\begin{samepage}
\problemx{Solving parity game}
{An arena $\mathcal{G}=(V_0,V_1,E,\Omega)$, an initial vertex $v_I \in V$.}
{Does player $0$ have a winning strategy from $v_I$ in the parity game $\mathcal{G}$?\newline}
\end{samepage}

One important property of parity games is that of {\em memoryless determinacy}: for a parity game and an initial vertex $v_I \in V$, one of the players has a memoryless winning strategy from $v_I$ \cite{zielonka1998infinite}.

>>>>

We consider here a different approach to extending pushdown systems, by allowing them to test the stack content against precise values.
% Pushdown systems which can test the stack content against regular-expressions are equivalent to pushdown systems, so pushdown systems which can test the stack content against specific words are too.
We consider 
%instead
here checks against unspecified word values that can be instantiated one after the other,
providing the pushdown systems the ability to
store information on some stack contents for later comparisons.

@@

\subsection{Games}

% Games represent a model for interaction between decision makers. We study so-called turn-based games in extensive form with perfect information where the players have to make decisions again and again.
\iffalse
After its foundation by von Neumann and Morgenstern [vNM44] in the
1940s, game theory has quickly become a valuable tool for modelling the
interaction of several agents with individual and possibly conflicting objec-
tives. It has applications in many scientific fields such as economy, sociology,
biology, logic and computer science. Whereas the games used in the former
three areas are usually finite in the sense that any play of these games ends
after a finite number of steps (in fact often after only a single step), the
games used in logic and computer science are in general infinite.

@article{copeland1945john,
  title={John von Neumann and Oskar Morgenstern, theory of games and economic behavior},
  author={Copeland, Arthur H},
  journal={Bulletin of the American Mathematical Society},
  volume={51},
  number={7},
  pages={498--504},
  year={1945},
  publisher={American Mathematical Society}
}

More precisely, an infinite game is played for an infinite number of
rounds. In each round, one player chooses an action depending on the se-
quence of actions chosen in previous rounds. After infinitely many rounds,
an infinite sequence of actions has emerged, called the outcome of the game.
Each player receives a payoff determined by a payoff function mapping each
possible outcome to a real number in the interval [0, 1]. We will concen-
trate on games where the possible payoffs are just 0 and 1, in which case a
winning condition, given as an abstract set of possible outcomes, is used to
determine whether a player wins (payoff 0) or loses (payoff 1) an outcome.
In logic, games are used to evaluate a logical formula in a mathematical
structure and to compare mathematical structures with respect to their
(in)distinguishability by a certain logic. Games of the former kind are called
model checking games; games of the latter kind are called model comparison
games. Whereas model checking and model comparison games for plain
modal or first-order logic are finite games, infinite games arise as model
checking games for fixed point logics and as model comparison games for
infinitary logics.
In computer science, games are used for the verification and synthesis
of reactive systems, i.e. systems that interact with their environment. As
soon as the systems under consideration do not necessarily terminate, these
games become infinite. For example, the behaviour of an operating system
can be understood as an infinite game between the system and its users.

So far, the games used in logic and computer science have usually been
two-player zero-sum games, i.e. games with only two players where one
player wins if and only if the other player loses. For example, in a model
checking game, one player wants to show that the given formula holds in the
given structure whereas the other player wants to show that this is not the
case. Analogously, in a game used for verification, one player wants to show
that the given system is able to react against its environment in such a way
that the resulting execution path of the system fulfils the given specification
whereas the other player wants to show that this is not the case.

The research on infinite games in theoretical computer science is based on a
mixture of several motivations:
1. The beautiful classical theory of infinite games, as developed in descriptive
set theory, lacks an algorithmic content. Such an algorithmic orientation is
provided by the automata theoretic approach to infinite games, originating
in work of Church, B ̈uchi, McNaughton, and Rabin about fourty years ago.
2. Determinacy results for infinite games are closely related to complementation
results for logics and automata; thus, infinite games help to analyze logical
theories (the most prominent being the monadic theory S2S of two successor
functions, see e.g. [Th97]).
3. Games are a natural model of reactive computation, and infinite games are
thus a faithful representation of nonterminating reactive systems (for which
control problems can be solved in terms of providing winning strategies).
4. The model checking problem for logics like the μ-calculus can be formulated
as the question to determine the winner of an infinite game.
The purpose of this tutorial is to explain the core of the algorithmic (and
automata theoretic) theory of infinite games. In the present extended abstract,
only some key notions are explained, without technical details and proofs. In the
tutorial, more will be said about topics of current research as listed in Section 6
of this abstract.

@incollection{thomas1997languages,
  title={Languages, automata, and logic},
  author={Thomas, Wolfgang},
  booktitle={Handbook of formal languages},
  pages={389--455},
  year={1997},
  publisher={Springer}
}

A game is composed of an arena and a winning condition.
\iffalse
\mh{Two-player games oﬀer a very convenient framework to represent
interaction of a program with some (possibly hostile) environment. In this approach, the ﬁrst player represents the program while the second player simulates
the environment. A winning strategy expresses a property that must hold whatever the environment does, thus ﬁnding a winning strategy for the ﬁrst player
allows to synthesize a controller that restricts the program and ensures that the
property expressed by the winning condition always holds \cite{arnold2003games}.}
\fi
We will ﬁrst study
arenas and
 then introduce common winning conditions that will be of interest to us.

 \subsubsection{Arenas}

\begin{samepage}
An {\em arena} is a tuple $A=(S_0, S_1,{\rightarrow})$  which is composed of
\begin{itemize}
	\item two  disjoint  sets of configurations, $S_0$ and $S_1$, %respectively associated with player $0$ and $1$, 
	whose disjoint union $S_0 \cup S_1$ we denote by $S$, and
	\item a relation of %labeled 
		transitions
	 ${\rightarrow} \subseteq S \times %\Lambda \times 
	 				S$. \\
\end{itemize}
\end{samepage}

\noindent
Note that $(S,{\rightarrow})$, where $S=S_0 \uplus S_1$, forms
a %labeled 
transition system, and, in particular, given a 
%labeled 
transition system,
one needs only to provide 
a partition of the set of configurations $S$ into two sets $S_0$ and $S_1$ to obtain an arena. \\
% We require that the out-degree of each configurations is at least one. This ensures that every finite path in (V, E) can be prolonged.

% \mh{I dont think I want to disallow dead-ends from the very beginning, means every time you define some arena you have to prove it has no dead-ends, seem tedious and unecessary.}

\noindent
The games we are interested in are played by two players, called {\em player $0$} and {\em player $1$}. We will often write player $i$ to denote a general player for $i \in \{0,1 \}$,
and we will call player $1-i$ its {\em opponent}.

% Observe that there is no restriction on the number of the successors of a configuration in an arena. Also, we don’t require that (V, E) is a bipartite graph with corresponding partition {V 0 , V 1 }.

\iffalse
\noindent
In the case where $\Lambda$ is a singleton $\{ a\}$, we will abuse notation
and denote the arena by
$(S_0, S_1,{\rightarrow},\Omega)$, and we will see $(S,{\rightarrow})$
as an unlabeled transition system.
\fi

 \subsubsection{Plays}

\noindent
%
%\noindent
A {\em play} in an arena $A=(S_0,S_1,\rightarrow)$ is a path in $(S,{\rightarrow})$ that is {\em maximal} in the following sense: it is either infinite or finite and if it is finite, then the target configuration is a dead end. 
A {\em partial play} in an arena $A=(S_0,S_1,\rightarrow)$ is a 
% sequence of configurations $s_{init} = s_0 s_1 \ldots \in S^\infty$ connected by the transition relation, i.e. such that $(v_i, v_{i+1}) \in T$.
prefix of a play in $A$.\\
%finite path in $(S,{\rightarrow})$.\\

\noindent
Essentially plays can be seen as this: the two players, player $0$ and player $1$, move along the labeled transition system, taking turns either inﬁnitely often or until a dead end is reached.

% \subsubsection{Strategies}
\subsection{Strategies}

% Before deﬁning formally what it means for a player to win a game, we need to introduce the notion of strategy.

A {\em strategy} for player $i \in \{ 0,1\}$ is a function $\sigma_i : S^* S_i \to S$ such
that $\sigma_i(ws)$ is a successor of $s$.
%
A strategy is called {\em memoryless} if its output only depends on the final configuration of the sequence $s \in S_i$, i.e. if $\sigma_i(ws) = \sigma_i(w's)$ for all $w,w' \in S^*$. Thus, a memoryless strategy for player $i$ can be written  as a function $\sigma_i: S_i \to S$.

A partial play $\pi = s_0 \rightarrow s_1 \rightarrow \ldots \rightarrow s_n$ is {\em consistent} with a strategy $\sigma_i$ if sequences of configurations that end in $S_i$ along this play have all successors according to the strategy, i.e.
if for all $j\in[0,n-1]$, for all element $v_j \in V_i$ in the sequence $\pi$,
$v_{j+1} = \sigma_i(v_j)$.
 Given a configuration $s$,   a strategy $\sigma_0$ for player $0$ and a strategy $\sigma_1$ for player $1$, the play starting with $s$ that is consistent with both strategies is unique and is called the {\em resulting play} of $\sigma_0$ and $\sigma_1$ starting from $s$. It is 
denoted by $\pi(s, \sigma_0, \sigma_1)$.\\

%\noindent
%Note that the above definitions are essentially the same for finite and infinite arenas.

 \subsubsection{Winning conditions}

Once we have provided players with an arena, it remains to define  properly what the winning condition is in order to define a game. Given an arena $A = (S_0, S_1, \rightarrow)$, a {\em winning condition}
%$\win \subseteq S^\infty$
$\win$
is a subset  
of the set of maximal plays for $A$. A {\em game} is a pair $\mathcal{G}= (A, \win)$ which is composed of an arena $A$ and a winning condition $\win$.
Although a winning condition is in general an infinite object we are interested in winning conditions 
that can finitely be described.

\iffalse
We consider here several types of winning conditions. 
Essentially the 
 mapping $\Omega$ of an arena serves to colour the configurations of the labeled transition system.  Winning conditions then
 consist in acceptance conditions on colour sequences.
 \fi

A strategy $\sigma_0$ for player $0$ from position $s \in S$ is a {\em winning strategy
 from $s$} for player $0$ if 
% no matter which
% strategy $\sigma_{1}$ the other player pursues,
% the play $\pi(s, \sigma_0, \sigma_1)$ is in $\win$.
every maximal play starting from $s$ and consistent with $\sigma_0$ is in $\win$.
On the other hand, a strategy $\sigma_1$ for player $1$ from position $s \in S$ is a {\em winning strategy
 from $s$} for player $1$ if 
% no matter which
% strategy $\sigma_{0}$ the other player pursues,
% the play $\pi(s, \sigma_0, \sigma_1)$ is not in $\win$.
every maximal play starting from $s$ and consistent with $\sigma_1$ is not in $\win$.

\iffalse
This definition of winning strategies is independent from the winning condition itself, and will also apply to %B\"uchi games and min-
parity games. To differentiate, we mention the winning conditions of a winning strategy from $s_{init}$.
\fi

% \paragraph{Reachability Games}
\paragraph{Reachability Games}

We first consider reachability games: 
the winning condition $\win_F(S)$ is given by a set of final configurations
$F \subseteq S$, and $\win_F(S)$ is the set of maximal plays in $S^* F S^\infty$. 
We simply write
$\win_F$ in case the set $S$ is obvious from context.
To indicate that the winning condition of a game is a reachability winning condition, we will speak of {\em reachability games}.

With regards to reachability games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem. \newline

\begin{samepage}
\problemx{Reachability game}
{A reachability game $\mathcal{G}= ((S_0, S_1,{\rightarrow}),\win_F)$, an initial configuration $s \in S_0 \cup S_1$.}
{Does player $0$ have a winning strategy from $s$ in $\mathcal{G}$?}
\end{samepage}

% \paragraph{Parity Games}
\paragraph{Parity Games}

Secondly, we consider min-parity games: 
the winning condition $\win_\Omega(S)$ is given by a {\em priority function} 
 $\Omega : S \to  [0, m]$.
An infinite path $ \pi = s_0 s_1 s_2 \ldots$ is in $\win_\Omega(S)$ if the smallest priority appearing infinitely often in
%this path
the sequence 
 is even, i.e.
 %$\min ~ \text{\textit{Inf}}( \Omega (r( \beta ))) $
if $\min( \{ \Omega(s) | ~ s \in \text{\textit{Inf}}( \pi) \})$ is even. 
% The winner of a finite play is the player whose opponent is unable to move, i.e. a play such that the last configuration of the play is in $S_{0}$ (resp. $S_1$) is player $1$ (resp. player $0$). 
A finite play is in $\win_\Omega(S)$ if its target configuration is in $S_1$.
We simply write
$\win_\Omega$ in case the set $S = S_0 \biguplus S_1$ is obvious from context.
To indicate that the winning condition of a game is a parity winning condition, we will speak of
 {\em parity games}.

With regards to parity games, we are going to concern ourselves with games over potentially infinitely large arenas, and we are interested in the following decision problem. \newline

\begin{samepage}
\problemx{Parity game}
{A parity game $\mathcal{G}= ((S_0, S_1,{\rightarrow}),\win_\Omega)$, an initial configuration $s \in S_0 \cup S_1$.}
{Does player $0$ have a winning strategy from $s$ in $\mathcal{G}$?}
\end{samepage}

%
\iffalse
Solving a parity game played on a finite graph means deciding, for a given starting position $v_{init}$, which of the two players has a winning strategy. It has been shown that this problem is in $\NP$ and 
co-$\NP$, more precisely $\UP$ and co-$\UP$. 
% Marcin Jurdziński (1998), "Deciding the winner in parity games is in UP∩ co-UP" (PDF), Information Processing Letters, Elsevier, 68 (3): 119–124, doi:10.1016/S0020-0190(98)00150-1
% @article{jurdzinski1998deciding,
%  title={Deciding the winner in parity games is in UP∩ co-UP},
%  author={Jurdzi{\'n}ski, Marcin},
%  journal={Information Processing Letters},
%  volume={68},
%  number={3},
%  pages={119--124},
%  year={1998},
%  publisher={Elsevier}
%}
\fi
%

% Muller condition: Inf() \in F

% Rabin condition :

% Streett Condition

% Rabin Chain condition

% max parity condition

% 1-winning  == reachability

\begin{center}
	\begin{figure}
			\hspace{3.81cm}
\includegraphics[width=0.4\textwidth]{figures/parity_game}
	\caption{An example of an arena. Circles  belong to player $0$, whereas diamonds belong to player $1$. Arrows represent transitions from a configuration to the next. The priority assigned by the priority mapping to each configuration is a number in $\{0,1,2,3\}$, next to the configuration it is assigned to. Note that we do not have a dead end in our example.}
	\label{example_parity_game}
	\end{figure}
\end{center}

\begin{example}
Let us consider the arena presented in  Figure~\ref{example_parity_game}. 
Configurations are partitioned into two sets: the circles, belonging to player $0$, and the diamonds,
belonging to player $1$. 
The priorities are $\{ 0,1,2,3\}$.
% The transitions and priority mappings may be derived from the picture.
% Note that we don't have a dead end in our example. 
Consider the game on the arena, where the winning condition is a parity condition.
Plays $\pi_1 = q_0 q_1 q_2 (q_4 q_7)^\omega$ and 
$\pi_2 = (q_0 q_1 q_2 q_4)^\omega$ are both winning for player $0$. Indeed, in $\pi_1$, 
the minimal priority encountered infinitely often is %that of $q_7$, that is, 
				$2$, 
whereas in $\pi_2$, it is %that of $q_2$, that is, 
			$0$. 
In both cases, it is an even number. On the other hand,
the plays $\pi_3 = (q_5 q_3 q_6)^\omega$ and $\pi_4=(q_5 q_6)^\omega$ are both winning for player $1$: the minimal priority encountered is odd, since it is $1$ for both.

One can take matters further: on this arena, player $0$ has a winning strategy for the parity game from $q_0$, $q_1$, $q_2$, $q_4$ and $q_7$, but not from $q_5$, $q_3$ nor $q_6$, where player $1$ has a winning strategy. 
\end{example}

% \mh{Say something abut positional determinacy}
One important property of parity games is that of {\em memoryless determinacy}: for a parity game
$\mathcal{G}= ((S_0,S_1,\rightarrow),\win_\Omega) $ and an initial configuration $s \in S_0 \cup S_1$, one of the players has a memoryless winning strategy from $s$ \cite{zielonka1998infinite}.

%\mh{
Of note is that several model checking problems %for pushdown automata 
can be expressed as decision problems for games: the most famous example is that the modal $\mu$-calculus model checking problem is polynomially equivalent to solving the parity game problem. 
%}
